---
title: "Class 10: Halloween Candy"
author: "Matin Nourmohammadi"
format: html
---

# 1. Importing candy data

```{r}
candy= read.csv('candy-data.txt',
                row.names = 1)

```

-   **Q1**. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

There are 85 types of candy.

-   **Q2**. How many fruity candy types are in the dataset?

```{r}
Fruity_candy_type= subset(candy, fruity==1)
nrow(Fruity_candy_type)
```

There are 38 fruity candies.

# 2. What is your favorate candy?

-   **Q3**. What is your favorite candy in the dataset and what is it's `winpercent` value?

```{r}
candy["Hershey's Kisses", 'winpercent']
```

-   **Q4**. What is the `winpercent` value for "Kit Kat"?

```{r}
candy["Kit Kat", 'winpercent']
```

-   **Q5**. What is the `winpercent` value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars", 'winpercent']
```

-   **Q6**. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

```{r}
#install.packages('skimr')
library(skimr)
skim(candy)
```

Winpercent seems to be different

-   **Q7**. What do you think a zero and one represent for the `candy$chocolate` column?

1 means true, 0 means false

-   **Q8**. Plot a histogram of `winpercent` values.

Histogram for `winpercent`

```{r}
hist(candy$winpercent)
```

-   **Q9**. Is the distribution of `winpercent` values symmetrical?

No, it is skewed to the left.

-   **Q10**. Is the center of the distribution above or below 50%?

Below 50%.

-   **Q11**. On average is chocolate candy higher or lower ranked than fruit candy?

We want to compare chocolate and fruity candy

```{r}
table(as.logical(candy$chocolate))
```

```{r}
length(candy$winpercent)
```

```{r}
candy$winpercent[ as.logical(candy$chocolate)]
```

```{r}
winpercent_chocolate <- candy$winpercent[ as.logical(candy$chocolate)]
mean(winpercent_chocolate)
```

For the fruity candies

```{r}
winpercent_fruity <-candy$winpercent[as.logical(candy$fruity)]
mean(winpercent_fruity)
```

Chocolate is higher than fruity candy.

-   **Q12**. Is this difference statistically significant?

Statistical test

```{r}
t.test(winpercent_chocolate, winpercent_fruity)
```

```{r}
boxplot(winpercent_chocolate, winpercent_fruity)
```

# 3. Overall Candy Rankings

-   **Q13**. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```

-   **Q14**. What are the top 5 all time favorite candy types out of this set?

```{r}
head(candy[order(-candy$winpercent),], n=5)
```

-   **Q15**. Make a first barplot of candy ranking based on `winpercent` values.

```{r}
library(ggplot2)

ggplot(candy) + aes(y=row.names(candy), x=winpercent) +
       geom_col(fill='orange')
```

-   **Q16**. This is quite ugly, use the `reorder()` function to get the bars sorted by `winpercent`?

```{r}
ggplot(candy) + aes(winpercent, reorder(rownames(candy),winpercent)) +
       geom_col(fill='orange')
```

Let's add other colors

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

-   **Q17**. What is the worst ranked chocolate candy?

Sixlets

-   **Q18**. What is the best ranked fruity candy?

starburst

# 4. Taking a look at pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

-   **Q19**. Which candy type is the highest ranked in terms of `winpercent` for the least money - i.e. offers the most bang for your buck?

```{r}
candy_win_more_than_80 <- candy[candy$winpercent>80,]
rownames(candy_win_more_than_80)[order(candy_win_more_than_80$pricepercent)]

```

Reese's miniatures

-   **Q20**. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

# 5. Exploring the correlation structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

-   **Q22**. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij, method = 'number')
```

Chocolate and fruity

-   **Q23**. Similarly, what two variables are most positively correlated?

Chocolate and winpercent

# 6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$rotation[,1])
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
library(plotly)
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

-   **Q24**. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

    fruity, hard, pluribus

    It does make sense since the general concept of candy does not include the three mentioned above.
